Dialect: A Compact Line Notation Based on SMILES

### 2022-03-18 manuscript.tex
Imports auto-generated bibliograph from Zotero. Fixes IUPAC element references. closes #97

\usepackage[backend=bibtex,sorting=none]{biblatex}
\addbibresource{manuscript}
\graphicspath{{assets/images/}}
\emergencystretch=1em

@@

SMILES has since been widely adopted. Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel;\cite{oBoyle:2011} RDKit;\cite{rdkit} Chemistry Development Kit;\cite{steinbeck2003} JChem Base;\cite{jchemBase} the Daylight Toolkit;\cite{daylightToolkit} and OEChem TK.\cite{oeChemTk} SMILES encodings can be found in many public-facing databases, including: PubChem;\cite{kim2016} ChEBI;\cite{degtyarenko2007} ZINC;\cite{irwin2005} ChEMBL;\cite{gaulton2012} and Wikipedia.\cite{wikipedia} Raw SMILES strings have been used extensively in machine learning applications.\cite{sousa2021} SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES,\cite{hanson2016} CurleySMILES,\cite{drefahl2011} BigSMILES,\cite{lin2019} and CXSMILES.\cite{cxsmiles}

@@

Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger. An accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson1987} Although the language described in that report shares many features with later descriptions, there are important differences as well.

@@

In 2019 IUPAC announced the SMILES+ initiative.\cite{smilesPlus} Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project. Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted.\cite{smilesPlusRepo}

@@

The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts. The FAIR Guiding Principles\cite{wilkinson2016} identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability. Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.

This situation has persisted long enough that the notion of a SMILES "standard" may not currently be feasible. A dozen or more slightly different implementations are now in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language "drift" while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of written documentation. The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use.\cite{daylightToolkit} For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned.\cite{depict} Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Dialect's system of representation is based on the \textit{molecular graph} concept.\cite{balaban1985} A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds"). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language\cite{murrayrust2011}, CDXML\cite{cdxml}, and Molfile\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

@@

Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}. The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC/IUPAP Working Party.\cite{iupac2016} When the Working Party authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Omitting the \ttt{isotope} property means that the isotopic composition is unspecified. When present, the lower bound on \ttt{isotope} equals one. This constraint leaves room for physically meaningless negative implied neutron count as in, for example, carbon-5. The lower bound for isotope applies even if the \ttt{element} attribute is not defined.

@@

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model).\cite{lewis1916} Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME occurs in a molecular graph when one or more equivalent representations exist, each one differing from the original only in the distribution of single and double bonds. DIME may be recognized as "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic2018,kerber2006}

@@

A filled DS can be emptied through a two-step process of \textit{deselection}. First, a perfect matching over the DS is found. Next, each matched edge is replaced by a double bond. Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds. A popular algorithm for matching, the Edmunds "blossom algorithm,"\cite{edmunds65} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes. Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.

@@

Not all tetracoordinate atoms will be \textit{stereocenters}. A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel.\cite{mislow1984} In the context of Dialect, a stereocenter must be tetracoordinate. Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

@@

Strings conform to an \textit{LL(1) grammar}. An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead. Additionally, LL(1) grammars expand the leftmost non-terminal first. This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers. LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR.\cite{parr2014} The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.

@@

To maximize compatibility with contemporary software, Dialect is based on the SMILES language. Given the widespread adoption of "SMILES," arriving at a specific definition of the language is surprisingly challenging. No single document completely specifies SMILES syntax and semantics. The first description of the SMILES language in the primary literature is contained within a 1988 article (referenced here as "the article").\cite{weininger:1988} The article, by its own admission, is incomplete: "... isomeric SMILES [defined in the book chapter as covering isotopism, conformation, and configuration (p. 91)] is not otherwise covered in this paper." (p. 34). Weininger later refined and expanded SMILES in a 2003 book chapter,\cite{weininger2008} referred to here as "the book chapter." A third major source of information is the Daylight Theory Manual (referred to here as "the manual"), a website operated by Daylight Chemical Information Systems, Inc.\cite{daylightTheory} Information about SMILES is also contained in various software packages and associated documentation, slide decks, websites, mailing lists, and private communications.

@@

The book chapter defines a fixed set of element symbols which even at the time of publication was obsolete (p. 83). The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{ohrstrom1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. These additions are one of the reasons Dialect can't be considered a strict subset of SMILES.

@@

The book chapter ascribes ambiguous meaning to detachments: "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other." (p. 88). The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

>>>>

\usepackage[backend=bibtex,sorting=none]{biblatex}
\addbibresource{manuscript}
\graphicspath{{assets/images/}}
\emergencystretch=1em
\DeclareUnicodeCharacter{2212}{-}

@@

SMILES has since been widely adopted. Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel;\cite{oBoyle:2011} RDKit;\cite{rdkit} Chemistry Development Kit;\cite{steinbeck:2003} JChem Base;\cite{jchemBase} the Daylight Toolkit;\cite{daylightToolkit} and OEChem TK.\cite{oetk} SMILES encodings can be found in many public-facing databases, including: PubChem;\cite{kim:2016} ChEBI;\cite{degtyarenko:2007} ZINC;\cite{irwin:2005} ChEMBL;\cite{gaulton:2012} and Wikipedia.\cite{wikipediaCAS} Raw SMILES strings have been used extensively in machine learning applications.\cite{sousa:2021} SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES,\cite{hanson:2016} CurleySMILES,\cite{drefahl:2011} BigSMILES,\cite{lin:2019} and CXSMILES.\cite{cxsmiles}

@@

Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger. An accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described in that report shares many features with later descriptions, there are important differences as well.

@@

In 2019 IUPAC announced the SMILES+ initiative.\cite{smilesPlusSpecification} Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project. Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted.\cite{smilesPlusGitHub}

@@

The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts. The FAIR Guiding Principles\cite{wilkinson:2016} identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability. Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.

This situation has persisted long enough that the notion of a SMILES "standard" may not currently be feasible. A dozen or more slightly different implementations are now in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language "drift" while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of written documentation. The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use.\cite{daylightToolkit} For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned.\cite{depictArchive} Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Dialect's system of representation is based on the \textit{molecular graph} concept.\cite{balaban:1985} A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds"). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language\cite{murray-rust:2011}, CDXML\cite{cdxml}, and Molfile\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

@@

Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}. The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC.\cite{periodicTable} When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Omitting the \ttt{isotope} property means that the isotopic composition is unspecified. When present, the lower bound on \ttt{isotope} equals one. This constraint leaves room for physically meaningless negative implied neutron count as in, for example, carbon-5. The lower bound for isotope applies even if the \ttt{element} attribute is not defined.

@@

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model).\cite{lewis:1916} Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME occurs in a molecular graph when one or more equivalent representations exist, each one differing from the original only in the distribution of single and double bonds. DIME may be recognized as "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic:2018,kerber:2006}

@@

A filled DS can be emptied through a two-step process of \textit{deselection}. First, a perfect matching over the DS is found. Next, each matched edge is replaced by a double bond. Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds. A popular algorithm for matching, the Edmunds "blossom algorithm,"\cite{edmonds:1965} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes. Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.

@@

Not all tetracoordinate atoms will be \textit{stereocenters}. A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel.\cite{mislow:1984} In the context of Dialect, a stereocenter must be tetracoordinate. Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

@@

Strings conform to an \textit{LL(1) grammar}. An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead. Additionally, LL(1) grammars expand the leftmost non-terminal first. This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers. LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR.\cite{parr:2014} The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.

@@

To maximize compatibility with contemporary software, Dialect is based on the SMILES language. Given the widespread adoption of "SMILES," arriving at a specific definition of the language is surprisingly challenging. No single document completely specifies SMILES syntax and semantics. The first description of the SMILES language in the primary literature is contained within a 1988 article (referenced here as "the article").\cite{weininger:1988} The article, by its own admission, is incomplete: "... isomeric SMILES [defined in the book chapter as covering isotopism, conformation, and configuration (p. 91)] is not otherwise covered in this paper." (p. 34). Weininger later refined and expanded SMILES in a 2003 book chapter,\cite{weininger:2008} referred to here as "the book chapter." A third major source of information is the Daylight Theory Manual (referred to here as "the manual"), a website operated by Daylight Chemical Information Systems, Inc.\cite{daylightTheory} Information about SMILES is also contained in various software packages and associated documentation, slide decks, websites, mailing lists, and private communications.

@@

The book chapter defines a fixed set of element symbols which even at the time of publication was obsolete (p. 83). The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. These additions are one of the reasons Dialect can't be considered a strict subset of SMILES.

@@

The book chapter ascribes ambiguous meaning to detachments: "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other." (p. 88). The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
